
```{r }
```

Reproducible Research: Peer Graded Assessment 1
===============================================
  Created by Prabhjot Singh on 25th June 2016

### Basic settings
```{r}

  # Always make code visible
  echo = TRUE  

  # Turn off scientific notations for numbers
  options(scipen = 1)  
```

### Loading and processing the data
```{r}
  # unzip the data file
  unzip("activity.zip")
  
  # read the data, with specific data types
  data <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
  
  data$month <- as.numeric(format(data$date, "%m"))
  
  # remove all na's
  cleanData <- na.omit(data)
  
  rownames(cleanData) <- 1:nrow(cleanData)
  
  # display head of clean data
  head(cleanData)
  
  # dimensions of clean data
  dim(cleanData)
  
  # Load the Library for plots
  library(ggplot2)
  
```


### What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

* Make a histogram of the total number of steps taken each day
```{r}
  # make a monthwise histogram of the number of steps
  ggplot(cleanData, aes(date, steps)) + 
    geom_bar(stat = "identity", colour = "blue", fill = "blue", width = 0.5) + 
    facet_grid(. ~ month, scales = "free") + 
    labs(title = "Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")

```

* Calculate and report the mean and median total number of steps taken per day

Mean total number of steps taken per day:

```{r}
    # Calcaute the mean
    totalSteps <- aggregate(cleanData$steps, list(Date = cleanData$date), FUN = "sum")$x
    mean(totalSteps)
    
```

Median total number of steps taken per day:

```{r}
  # Calcaute the median
  median(totalSteps)

```

### What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

avgSteps <- aggregate(cleanData$steps, list(interval = as.numeric(as.character(cleanData$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + 
  geom_line(color = "blue", size = 0.5) + 
  labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")

```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
  avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```

### Imputing missing values
* The total number of rows with NAs:
  
```{r}
  # Calculate and report the total number of missing values in the dataset
  sum(is.na(data))

```

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

My strategy would be to use the mean for that 5-minute interval to fill each NA value in the steps column.

* Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
  
  # My strategy - use the mean for that 5 minute interval
  
  newData <- data 
  for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
      newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
  }

  head(newData)

  # now check total number of missing values
  sum(is.na(newData))
  
```

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}

  ggplot(newData, aes(date, steps)) + 
    geom_bar(stat = "identity", colour = "blue", fill = "blue", width = 0.5) + 
    facet_grid(. ~ month, scales = "free") + 
    labs(title = "Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")

```

* Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean total number of steps taken per day:
```{r}
  
  newTotalSteps <- aggregate(newData$steps,
                          list(Date = newData$date),
                          FUN = "sum")$x
  newMean <- mean(newTotalSteps)
  newMean
  
```
Median total number of steps taken per day:
```{r}

  newMedian <- median(newTotalSteps)
  newMedian

```

Compare them with the two before imputing missing data:

```{r}

  oldMean <- mean(totalSteps)
  oldMedian <- median(totalSteps)
  newMean - oldMean
  newMedian - oldMedian

```
So, after imputing the missing data, the new mean of total steps taken per day is the same as that of the old mean; the new median of total steps taken per day is greater than that of the old median.

### Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}

  head(newData)
  newData$weekdays <- factor(format(newData$date, "%A"))
  levels(newData$weekdays)
  levels(newData$weekdays) <- list(weekday = c("Monday", 
                                               "Tuesday",
                                               "Wednesday", 
                                               "Thursday", 
                                               "Friday"),
                                   weekend = c("Saturday", 
                                               "Sunday"))
  levels(newData$weekdays)
  table(newData$weekdays)

```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
  avgSteps <- aggregate(newData$steps, 
                        list(interval = as.numeric(as.character(newData$interval)), 
                             weekdays = newData$weekdays),
                        FUN = "mean")

  names(avgSteps)[3] <- "meanOfSteps"

  library(lattice)
  
  xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
         layout = c(1, 2), type = "l", 
         xlab = "Interval", ylab = "Number of steps",
         col="blue")
  
```
```
```
```

